---
title: "Using `{sf}` operations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using `{sf}` operations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 300  
)
```

Because `{densityarea}` can return `{sf}` polygons, this can allow you to use its functionality (see [this cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/sf.pdf) for some examples).

```{r setup}
library(densityarea)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)

library(sf)

library(ggplot2)
```

## Density polygons as simple features

As a first example, we'll estimate how much different data clusters overlap in the `s01` dataset.

```{r}
data(s01)

s01 |> 
  mutate(lF1 = -log(F1),
         lF2 = -log(F2)) ->
  s01
```

We'll focus on the vowels `iy`, `ey`, `o` and `oh` which correspond to the following lexical classes:

| vowel label | lexical class         |
|-------------|-----------------------|
| `iy`        | [Fleece]{.smallcaps}  |
| `ey`        | [Face]{.smallcaps}    |
| `o`         | [Lot]{.smallcaps}     |
| `oh`        | [Thought]{.smallcaps} |

```{r}
s01 |> 
  filter(
    plt_vclass %in% c("iy", 
                      "ey", 
                      "o", 
                      "oh")
  ) ->
  vowel_subset
```

Within this subset of vowel categories, we'll get the 80% probability density estimate as `sf::st_polygon()`s.

```{r}
vowel_subset |> 
  group_by(plt_vclass) |> 
  reframe(
    density_polygons(lF2, 
                     lF1, 
                     probs = 0.8,
                     as_sf = TRUE)
  ) |> 
  st_sf()->
  vowel_polygons

vowel_polygons
```

We can plot these directly by using the `sf::geom_sf()` geom for ggplot2.

```{r fig.width=5, fig.height=5, out.width="80%", fig.align='center'}
ggplot(vowel_polygons) +
  geom_sf(
    aes(fill = plt_vclass),
    alpha = 0.6
  ) +
    scale_fill_brewer(palette = "Dark2")
```

### Initial `{sf}` operations

All of the `{sf}` operations for geometries are available to use on `vowel_polygons`. For example, we can get the area of each polygon, with `sf::st_area()` and use it in plotting.

```{r}
vowel_polygons |> 
  mutate(
    area = st_area(geometry)
  ) ->
  vowel_polygons
```

```{r fig.width=5, fig.height=5, out.width="80%", fig.align='center'}
ggplot(vowel_polygons) +
  geom_sf(
    aes(fill = area),
    alpha = 0.6
  )+
  scale_fill_viridis_c()
```

Or, we can get the polygon centroids and plot them.

```{r fig.width=5, fig.height=5, out.width="80%", fig.align='center'}
vowel_polygons |> 
  st_centroid() |> 
  ggplot()+
    geom_sf_label(
      aes(label = plt_vclass,
          color = plt_vclass,
          size = area)
    )+
    scale_color_brewer(palette = "Dark2")+
    coord_fixed()
```

### Getting overlaps

To use the density polygons like "cookie cutters" on each other, we need to use `st_intersections()`.

```{r}
vowel_polygons |> 
  st_intersection() -> 
  vowel_intersections

vowel_intersections
```

This data frame contains a polygon for each unique intersection of the input polygons, with a new `n.overlaps` column.

```{r fig.width=5, fig.height=5, out.width="80%", fig.align='center'}
ggplot(vowel_intersections) +
  geom_sf(
    aes(fill = n.overlaps)
  )+
  scale_fill_viridis_c()
```

The labels of the new overlapping regions aren't very informative, but we can create some new labels by using the indices in the `origins` column.

```{r}
new_label <- function(indices, labels){
  str_c(labels[indices],
        collapse = "~")
}
```

```{r}
vowel_intersections |> 
  mutate(
    groups = map_chr(
      origins,
      .f = new_label,
      labels = vowel_polygons$plt_vclass
    ) 
  ) |> 
  relocate(groups, .after = plt_vclass)->
  vowel_intersections

vowel_intersections
```

```{r fig.width=5, fig.height=5, out.width="80%", fig.align='center'}
ggplot(vowel_intersections)+
  geom_sf(
    aes(fill = groups)
  )+
  scale_fill_brewer(palette = "Dark2")
```

We can also calculate the areas of these new polygons, and compare them to the original areas (which have been preserved in `areas`.

```{r fig.width=5, fig.height=3, out.width="80%", fig.align='center'}
vowel_intersections |> 
  mutate(
    group_area = st_area(geometry),
    overlapped_proportion = 1-(group_area/area)
  ) |> 
  filter(n.overlaps == 1) |> 
  ggplot(
    aes(plt_vclass, overlapped_proportion)
  )+
    geom_col()+
    ylim(0,1)
```

## Spatial filters and joins

There are also a number of spatial filters and merges that can be used interestingly if the original data points are also converted to sf objects.

```{r}
library(sfheaders)
library(forcats)
```

```{r}
s01 |> 
  sfheaders::sf_point(
    x = "lF2",
    y = "lF1",
    keep = TRUE
  ) ->
  s01_sf

s01_sf
```

```{r fig.width=5, fig.height=5, out.width="80%", fig.align='center'}
s01_sf |> 
  ggplot()+
    geom_sf()
```

Next, we can get the density polygon for a single vowel,.

```{r}
s01 |> 
  filter(plt_vclass == "iy") |> 
  reframe(
    density_polygons(lF2, lF1, probs = 0.8, as_sf =T)
  ) |> 
  st_sf()->
  iy_sf
```

### Spatial filter

Let's get all of the points in `s01_sf` that are "covered by" the `iy_sf` polygon.

```{r}
s01_sf |> 
  st_filter(
    iy_sf,
    .predicate = st_covered_by
  )->
  covered_by_iy
```

```{r fig.width=5, fig.height=5, out.width="80%", fig.align='center'}
covered_by_iy |> 
  mutate(plt_vclass = plt_vclass |> 
           fct_infreq() |> 
           fct_lump_n(5)) |> 
  ggplot()+
    geom_sf(data = iy_sf)+
    geom_sf(aes(color = plt_vclass))+
    scale_color_brewer(palette = "Dark2")
```

Obviously, the 80% probability density area for `iy` is not a homogeneous region.

```{r fig.width=5, fig.height=3, out.width="80%", fig.align='center'}
covered_by_iy |> 
  mutate(plt_vclass = plt_vclass |> 
           fct_infreq() |> 
           fct_lump_n(5)) |> 
  count(plt_vclass) |> 
  ggplot(aes(plt_vclass, n))+
    geom_col(
      aes(fill = plt_vclass)
    )+
    scale_fill_brewer(palette = "Dark2")
```

### Spatial joins

For a spatial join, let's see which vowel category a random vowel token is close to.

```{r}
set.seed(100)
s01_sf |> 
  slice_sample(n = 1)->
  rand_vowel

rand_vowel
```

Then, we'll get density polygons at a few different probability points for all vowels.

```{r}
s01 |> 
  group_by(plt_vclass) |> 
  reframe(
    density_polygons(
      lF1,
      lF2,
      probs = ppoints(3),
      as_sf = T
    )
  )
```
